Identification of Direct Cherenkov Pixels using Boosted Decision Trees

### 2016-07-01 report/draft1.tex
Backup 01/07/16

\documentclass{article}
\usepackage{graphicx}
\graphicspath{ {graphs/} }
\usepackage{geometry}
\geometry{a4paper, portrait, margin=1.0in}
\usepackage{gensymb}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{pgfplotstable, booktabs}

@@

\section{Introduction}
Cosmic Rays are abundant in our galaxy, and enter the Earth's upper atmosphere from every direction. Their flux is governed by a well-defined energy power law, which is almost unchanged across more than 10 orders of magnitude \cite{Blasi:2013rva}. Cosmic Rays are in fact not rays but rather charged particles, primarily protons, which have been accelerated by various galactic or stellar processes. The exact cosmic ray composition is energy dependent, and provides information regarding the source of cosmic rays. Heavy cosmic rays, the most common of which is iron, vary substantially in relative abundance for different energy scales.

Being highly-energetic charged particles, Cosmic rays will often travel through the atmosphere faster than the local speed of light. Consequently, many Cosmic Rays will emit Cherenkov light over substantial distances, and this light will be visible on the ground. There are numerous Telescope Arrays which image the Cherenkov Light emitted by Cosmic Rays in the atmosphere, including the HESS, MAGIC and VERITAS Experiments. Although the number of emitted photons is comparatively small, the Cherenkov light is still brighter than the night sky background because the telescopes are able to capture images with an exceptionally short recording time, equal to 80ns for the HESS array \cite{Funk:2004ie}. The Cherenkov photons all arrive almost simultaneously, so have a high intensity over the short time window during which the image is formed.

All telescope arrays rely on Hillas Analysis with extracted parameters from each of the camera images being used to reconstruct the events, but heavy atmospheric blurring means that charge resolution is very poor. For Iron Nucleus events, we would expect to reconstruct \[Z \approx 26 \pm 5 \] with a core position resolution of roughly $d \approx 20 m $ \cite{hess07}. The Cosmic Rays imaged by these telescopes have energies between $13-200 $TeV, and at present, no study of the relative abundance of different cosmic ray elemental abundances exists at these energies. If the telescope arrays could be used to spectroscopically analyse the Cosmic rays, it could provide important clues regarding the mechanism of Cosmic Ray formation and propagation in the galaxy. However, current charge resolution from Hillas Analysis is not small enough to undertake such a study. The principal goal of this thesis is to enable spectroscopic analysis with Cherenkov Telescope arrays, by improving the charge resolution of reconstructed events.

A theoretical study by Kieda in 2001 \cite{kieda01} suggested that for a core position resolution of $d \approx 5 $ m, we could expect to see a charge resolution of $ \sigma_{Z} \approx 1 $ for elements of $Z = 20$ or higher. In this case the core position resolution would be the limiting value. Thus, if the LPD method can achieve this core position resolution, the precision will be sufficient to extract the abundances of the different Cosmic Ray Elements.

@@

To understand why an LPD exists, it is important to recall that Cherenkov emission begins when a charged particle passes the local speed of light in a medium.  We know that the refractive index of air is proportional to the density of air. Thus, the minimum velocity needed for Cherenkov Emission is lower for low atmospheric altitudes, because the air is denser. If the velocity is sufficient, we will observe photon emission from the particle at the Cherenkov Angle $\theta_{c}$.

\[ \cos(\theta_{c}) = \frac{1}{\eta \beta}  \Longrightarrow \beta_{min} \geq \frac{1}{\eta}\]

@@

The salient point is that there is a clear characteristic photon distribution that, for energies above 25 TeV at least, is almost constant except for a varying endpoint. However, we will not be able to directly measure this LPD. Instead, a telescope images containing air showers must be analysed, and from these images, we can infer the quantity of DC light present amongst a large degree of background. Extensive simulations of Camera images will be performed, in order to quantify the resultant LPD extracted from the telescope images. After determining a likely experimental LPD, Monte Carlo events can be simulated, and then reconstructed.

To reconstruct an event, we will need to find the \textquoteleft Core position' of the Cosmic Ray, which is defined as the position on the ground through which the Cosmic Ray would pass if it continued in a straight line to the ground. In addition, we will need to determine the Energy, first interaction height, and the charge. We would normally require at least five data points in order to reconstruct these variables. However, it is possible to use the non-observation of light in one telescope to constrain the core position. Thus, we require that an  event was be seen by at least four telescopes for a five-telescope array such as HESS.

\section{DC Pixel Identification}
\textbf{Cosmic Rays will emit Direct Cherenkov (DC) light in the upper atmosphere, before generating an Extended Air Shower (EAS) through interaction with the lower atmosphere. }

In telescope images, the DC light is usually concentrated in a single  \textquoteleft DC pixel'. Identifying this pixel is challenging, because the brighter EAS Cherenkov light background often overlaps with the DC pixel. In order to apply the LPD method to data, we must first identify the DC pixel in a shower image, and determine the number of DC photoelectrons present.

We define the variable $Q_{DC} = \frac{Intensity_{N.N.max}}{Intensity}$ as the ratio of the largest neighbouring pixel intensity to the intensity of a given pixel. In previous experiments, the DC pixel candidate was identified by applying a number of cuts to pixels in an image \cite{hess07}, and from the subset of pixels passing the specified cuts, selecting the pixel with the smallest $Q_{DC}$ as the \textquoteleft DC candidate'. Due to the low pass rate for cuts, we obtain a small low-contamination dataset, while the majority of telescope images are left without a DC pixel candidate. An improved method would aim to increase the number of correctly identified DC pixels, while still enabling cuts which discriminate well between correctly and incorrectly identified DC pixels.

@@

\subsection{Image Simulation}
The full simulation of air showers was performed using the CORSIKA package, with a standard atmospheric profile derived from measurements conducted at the HESS site in Namibia. In total, 5000 training events and a further 10000 testing events were simulated. The simulated particles were $Fe^{56}$, within the Energy Range of $35-135$ TeV and a spectrum $\phi(E) \propto E^{-2.7}$. For each set of simulated event, 4 unique random number seeds were used to generate the shower. An altitude of 1800m was assumed, again corresponding to the HESS site. The simulated zenith angle ranged from $0\degree < \theta < 2 \degree$, while the simulated azimuth angle ranges from $-2\degree < \phi < 2 \degree$. The four smaller HESS-phase-1 telescopes were arranged in a cross along the x/y axis with the larger HESS-phase-2 \textquoteleft CT5' telescope placed at the center. The length of each cross arm was $85m$. The simulated target region of the cores was chosen to be a square centered on CT5, with each 300m-long side bisecting the x/y axis.Due to hardware differences between CT5 and the original HESS 1 telescopes, we analyse HESS1 and HESS2 images separately. To determine the true class of each pixel, a simulation was initially run with an energy cut of 10 PeV on all muons and electrons. Because this cut exceeded the primary particle energy, neither daughter muons and electrons, nor the photons they would have emitted, were simulated. Thus the hadronic Cherenkov Light from the primary particle and daughter fragments, but not the EAS light, was present in the camera image. A second identical \textquoteleft EAS Simulation' was run including the same random seeds, but without the energy cut on muons and electrons. This gave a complete EAS image including identical DC light.

@@

@@

Within CORSIKA, a square simulation of 300 x 300m was simulated. We can consider an identical $90000 m^{2}$ target region for core position simulation. The HESS telescope additionally has a solid and field of view of 5 degrees (0.09 rad),meaning an expected telescope array Iron Flux of $F(35-135TeV) = 0.22 s^{-1}$. On the basis of efficiency calculations in DC pixel identification, it is clear that 2.2\% of Iron Ray events will produce high-multiplicity DC pixels accepted by the BDT. Thus, we will ultimately obtain an hourly flux of $F(35-135TeV) = 4.84 \times 10^{-3} s^{-1} = 17.4 h^{-1}$ high-multiplicity Iron events.

The full 5-telescope HESS array has been in operation since 2012 \cite{hessCT5}, and has thus had time to collect several years of data. Due to the requirement of night operation under favourable weather conditions, the HESS phase 2 experiment has collected approximately 5000 hours of data. On the basis of our high-multiplicity flux rate, we can assume that a rough expected event count would be roughly 7900 high-multiplicity iron events. The Monte Carlo simulations will be run with approximately this number of events.

To accurately model the Cosmic Rays, the Energy Power Law must be simulated. To generate the random Energy value En(R) following the power law, a uniformly distributed random number R is generated. The value of R represents a random fraction of the total simulated flux $ F(R) = R \times F(35-135TeV) $, and can thus range from 0-1. Having determined the Integrated Flux corresponding to the random number, the Energy corresponding to this flux can be calculated. The energy E is defined as the lower bound of the integral which would produce the chosen integrated flux F. Thus R=0 corresponds to an Energy of 135TeV yielding no Integrated Flux, while R=1 would correspond to an Energy of 35TeV and the full Integrated Flux. THus the simulated Cosmic Rays will obey the power law, with high-energy events being realistically suppressed.

\[ F(R) = F(En(R)-135TeV) = \int_{En(R)}^{135} \phi_{0} E^{-2.76} dE =\frac{\phi_{0}}{1.76}[En(R)^{-1.76} - 135^{-1.76}]= R \times F(35-135TeV)\] \[\Longrightarrow  En(R) = (\frac{1.76\times R \times F(35-135TeV)}{\phi_{0}} +135^{-1.76})^{\frac{-1}{1.76}} \]

RATES!!!

@@

\begin{table}[h!]
  \centering
  \caption{Fractional error from $True_{DC}$ for all high-multiplicity test pixels}
  \label{tab:altlpderror}
  \begin{tabular}{ccc}
    \toprule
    & HESS1 & HESS2\\
    \midrule
    $\sigma_{rgr}$ & 0.32 & 0.29\\ 
    $\sigma_{rejected}$ & 0.41 & 0.63\\ 
    \bottomrule
  \end{tabular}
\end{table}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{DCcounterrorhess1rgrbigtestdata}
\caption{The difference between true and reconstructed DC signal for events passing the cuts is shown in the top left. In the top right the same is shown for events which were rejected. The true DC signals for events which passed the cuts is plotted against the reconstructed signals in the lower left, and for rejected events in the lower right. As in \ref{fig:cutdistribution}, in all plots a green event is one in which the DC pixel has been correctly identified, while a red event is one that has been incorrectly identified.}
\label{fig:dcdiff}
\end{center}
\end{figure}
\section{Bibliography}
\bibliographystyle{unsrt}
\bibliography{report}
\section{Appendix}
\subsection{Atmospheric Data Table for CORSIKA (Windhoek, Namibia)}
\pgfplotstabletypeset{data/atmprof10.dat}
\end{document}

>>>>

\documentclass[11pt]{article}
\usepackage{graphicx}
\graphicspath{ {graphs/} }
\usepackage{geometry}
\geometry{a4paper, portrait, margin=1.0in}
\usepackage{gensymb}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{pgfplotstable, booktabs}

@@

\section{Introduction}
Cosmic Rays are abundant in our solar system, entering the Earth's upper atmosphere from every direction. Cosmic Rays are in fact not rays but rather charged particles, primarily protons, which have been accelerated by various galactic or stellar processes. Their flux is governed by a well-defined energy power law, which is almost unchanged across more than 10 orders of magnitude \cite{Blasi:2013rva}.  The exact Cosmic Ray composition is energy dependent, and provides information regarding their sources. Heavy Cosmic Rays, the most common of which is iron, vary substantially in relative abundance for different energy scales.

Being highly-energetic charged particles, Cosmic Rays will often travel through the atmosphere faster than the local speed of light. Consequently, many will emit substantial Cherenkov light, and this light will often be visible on the ground. There are numerous Telescope Arrays which image the Cherenkov Light emitted by Cosmic Rays in the atmosphere, including the HESS, MAGIC and VERITAS Experiments. Although the number of emitted photons is comparatively small, the Cherenkov light is still distinguishable from the night sky background because the telescopes are able to capture images with an exceptionally short recording time. The Cherenkov photons all arrive almost simultaneously, so have a high intensity during the short time window when the image is formed. For the HESS array, this window is equal to 80ns \cite{Funk:2004ie}.

All such telescope arrays currently rely on Hillas Analysis to fully reconstruct the events, including the  charge Z of the Cosmic Ray. Hillas Analysis relies on extracted parameters from each of the camera images, but heavy atmospheric blurring of these images means that charge resolution is very poor. For Iron Nucleus events, we would expect to reconstruct \[Z \approx 26 \pm 5 \] with a core position resolution of roughly $d \approx 20 m $ \cite{hess07}. The Cosmic Rays imaged by these telescopes have energies in TeV scale, and at present, no study of the relative abundance of different cosmic ray elemental abundances exists at these energies. If the telescope arrays could be used to spectroscopically analyse the Cosmic Rays, it could provide important clues regarding the mechanism of Cosmic Ray formation and propagation in the galaxy. However, current charge resolution from Hillas Analysis is not small enough to undertake such a study. The principal goal of this thesis is to enable spectroscopic analysis with Cherenkov Telescope arrays, by improving the charge resolution of reconstructed events.

A theoretical study by Kieda in 2001 \cite{kieda01} suggested that, with a core position resolution of $d \approx 5 $ m, we could expect to see a charge resolution of $ \sigma_{Z} \approx 1 $ for elements of $Z = 20$ or higher. In this case the core position resolution would be the limiting value. Thus, if the LPD method can achieve this core position resolution, the precision will be sufficient to extract the abundances of the different Cosmic Ray Elements.

@@

To fully reconstruct an event, we will need to find the charge, energy and the core position coordinates. In addition, we will need to determine the height of the first atmospheric interaction. We would normally require at least five data points in order to reconstruct these variables. However, it is possible to use the non-observation of light in one telescope to constrain the core position. Thus, we require that an event was be seen by at least four telescopes for a five-telescope array such as HESS.

\section{DC Pixel Identification}
Cosmic Rays will emit Direct Cherenkov (DC) light in the upper atmosphere, before generating an Extended Air Shower (EAS) through interaction with the lower atmosphere. In telescope images, the DC light is usually concentrated in a single  \textquoteleft DC pixel'. Identifying this pixel is challenging, because the brighter EAS Cherenkov light background often overlaps with the DC pixel. In order to apply the LPD method to data, we must first identify the DC pixel in a shower image, and determine the number of DC photoelectrons present.

\subsection{Theoretical LPD}
To understand why DC light has a characteristic distribution, it is important to recall that Cherenkov emission begins when a charged particle passes the local speed of light in a medium.  We know that the refractive index of air is proportional to the density of air. Thus, the minimum velocity needed for Cherenkov Emission is lower for low atmospheric altitudes, because the air is denser. If the velocity is sufficient, we will observe photon emission from the particle at the Cherenkov Angle $\theta_{c}$.

\[ \cos(\theta_{c}) = \frac{1}{\eta \beta}  \Longrightarrow \beta \geq \frac{1}{\eta}\]

@@

We see that there is a clear characteristic photon distribution which, for energies above 25 TeV, is almost constant except for a varying endpoint. However, we will not be able to directly measure this LPD. Instead, a telescope images containing air showers must be analysed, and from these images, we can infer the quantity of DC light present amongst a large degree of background. Extensive simulations of Camera images will be performed, in order to quantify the resultant LPD extracted from the telescope images. After determining a likely experimental LPD, Monte Carlo events can be simulated, and then reconstructed.

\subsection{Image Simulation}
We can find the DC pixel by defining the variable $Q_{DC} = \frac{Intensity_{N.N.max}}{Intensity}$ as the ratio of the largest neighbouring pixel intensity to the intensity of a given pixel. In previous experiments, the DC pixel candidate was identified by applying a number of cuts to pixels in an image \cite{hess07}, and from the subset of pixels passing the specified cuts, selecting the pixel with the smallest $Q_{DC}$ as the \textquoteleft DC candidate'. Due to the low pass rate for cuts, we obtain a small low-contamination dataset, while the majority of telescope images are left without a DC pixel candidate. An improved method would aim to increase the number of correctly identified DC pixels, while still enabling cuts which discriminate well between correctly and incorrectly identified DC pixels.

@@

The full simulation of air showers was performed using the CORSIKA package with a standard atmospheric profile derived from measurements conducted at the HESS site in Namibia. This Atmospheric Profile is included in the Appendix. In total, 5000 training events and a further 10000 testing events were simulated. The simulated particles were $Fe^{56}$, within the Energy Range of $35-135$ TeV and a flux spectrum $\phi(E) \propto E^{-2.7}$. For each set of simulated event, 4 unique random number seeds were used to generate the shower. An altitude of 1800m was assumed, again corresponding to the HESS site. The simulated zenith angle ranged from $0\degree < \theta < 2 \degree$, while the simulated azimuth angle ranges from $-2\degree < \phi < 2 \degree$. The four smaller HESS-phase-1 telescopes were arranged in a cross along the x/y axis with the larger HESS-phase-2 \textquoteleft CT5' telescope placed at the center. The length of each cross arm was $85m$. The simulated target region of the cores was chosen to be a square centered on CT5, with each 300m-long side bisecting the x/y axis.Due to hardware differences between CT5 and the original HESS 1 telescopes, we analyse HESS1 and HESS2 images separately. To determine the true class of each pixel, a simulation was initially run with an energy cut of 10 PeV on all muons and electrons. Because this cut exceeded the primary particle energy, neither daughter muons and electrons, nor the photons they would have emitted, were simulated. Thus the hadronic Cherenkov Light from the primary particle and daughter fragments, but not the EAS light, was present in the camera image. A second identical \textquoteleft EAS Simulation' was run including the same random seeds, but without the energy cut on muons and electrons. This gave a complete EAS image including identical DC light.

@@

\subsection{Low-Energy Emission}
In the theoretical simulation of the LPD, we noted that the distribution in Figure \ref{fig:lpd} is only saturated in the region of around 35 TeV. Below this, the LPD is more heavily dependent on Energy, and thus our basic assumption of a characteristic equation will no longer be valid. To account for this, an additional simulation of 15000 Iron events was conducted in the energy range 5-135TeV. The BDT remained the same, and the same initial cuts were replicated as before. The same was also done for Protons in the same energy range. The results are summarised in Table \ref{tab:fullironacceptance}.

\begin{table}[h!]
  \centering
  \caption{Comparison of Proton and Iron Acceptance}
  \label{tab:fullironacceptance}
  \begin{tabular}{ccc}
    \toprule
    & HESS1  & HESS2 \\
    \midrule
    Proton Pixels Accepted after All Cuts(\%) & 0.21 & 0.19\\
    Sample Purity (\%) & 1.74 & 0.00 \\
    \midrule
    Iron Pixels Accepted after All Cuts(\%) & 0.17 & 0.11\\
    Sample Purity (\%) & 80.00 & 80.00 \\
    \bottomrule
  \end{tabular}
\end{table}

We find that, for Iron Events, the acceptance rate falls by a factor of approximately 15. The relative proportion of HESS 1 to HESS 2 acceptance is broadly maintained at a factor of roughly 1.5, although the sample purity falls somewhat. If we account for the differing intergrated flux rate, we can calculate the ratio of expected high-multiplicity event rate $\Gamma_{HM}(E)$ for the ranges 5-135 TeV and 35-135 TeV.

\[ \frac{\Gamma_{HM}(5-135)}{\Gamma_{HM}(35-135)} = \frac{5^{-1.7} - 135^{-1.7}}{35^{-1.7} - 135^{-1.7}} 
= 30.3 \]

Thus, if we have thirtyfold increase in event rate, but only one-fifteenth the acceptance rate, we would ultimately expect that we would have twice as many Iron Events if the full energy range of 5-135 TeV was included. However, the low-energy proton acceptance is much higher. Thus, were we to include events in this energy range, we would again retrieve a sample dominated by protons that will not be correctly reconstructed. The energy distribution of the Protons and Iron Events are shown in FIG. To remove this contamination from protons and non-saturated Iron events, we thus apply one additional Energy cut on the data sample. We simply require that the reconstructed Energy is greater than 30TeV. This will, assuming a rough energy resolution of around 15\% in reconstruction, remove the majority of these low-energy events. (WILL CONFIRM ONCE HISTOGRAMS ARE RETRIEVED).

@@

Within CORSIKA, a square simulation of 300 x 300m was simulated. We can consider an identical $90000 m^{2}$ target region for core position simulation. The HESS telescope additionally has a solid and field of view of 5 degrees (0.006 steradians),meaning an expected telescope array Iron Flux of $F(35-135TeV) = 0.0154 s^{-1}$. On the basis of efficiency calculations in DC pixel identification, it is clear that 2.2\% of Iron Ray events will produce high-multiplicity DC pixels accepted by the BDT. Thus, we will ultimately obtain an hourly flux of $F(35-135TeV) = 3.40 \times 10^{-4} s^{-1} = 1.22 h^{-1}$ high-multiplicity Iron events.

The full 5-telescope HESS array has been in operation since 2012 \cite{hessCT5}, and has thus had time to collect several years of data. Due to the requirement of night operation under favourable weather conditions, the HESS phase 2 experiment has collected approximately 5000 hours of data. On the basis of our high-multiplicity flux rate, we can assume that a rough expected event count would be roughly 6100 high-multiplicity iron events. The Monte Carlo simulations will be run with approximately this number of events.

To accurately model the Cosmic Rays, the Energy Power Law must be simulated. To generate the random Energy value En(R) following the power law, a uniformly distributed random number R is generated. The value of R represents a random fraction of the total simulated flux $ F(R) = R \times F(35-135TeV) $, and can thus range from 0-1. Having determined the Integrated Flux corresponding to the random number, the Energy corresponding to this flux can be calculated. The energy E is defined as the lower bound of the integral which would produce the chosen integrated flux F. Thus R=0 corresponds to an Energy of 135TeV yielding no Integrated Flux, while R=1 would correspond to an Energy of 35TeV and the full Integrated Flux. Thus the simulated Cosmic Rays will obey the power law, with high-energy events being realistically suppressed.

\[ F(R) = \int_{En(R)}^{135} \phi_{0} E^{-2.76} dE =\frac{\phi_{0}}{1.76}[En(R)^{-1.76} - 135^{-1.76}]= R \times F(35-135TeV)\] \[\Longrightarrow  En(R) = (\frac{1.76\times R \times F(35-135TeV)}{\phi_{0}} +135^{-1.76})^{\frac{-1}{1.76}} \]

@@

\begin{table}[h!]
  \centering
  \caption{Fractional error from $True_{DC}$ for all high-multiplicity test pixels}
  \label{tab:altlpderror}
  \begin{tabular}{ccc}
    \toprule
    & HESS1 & HESS2\\
    \midrule
    $\sigma_{rgr}$ & 0.32 & 0.29\\ 
    $\sigma_{rejected}$ & 0.41 & 0.63\\ 
    \bottomrule
  \end{tabular}
\end{table}
\section{Bibliography}
\bibliographystyle{unsrt}
\bibliography{report}
\section{Appendix}
\subsection{Atmospheric Data Table for CORSIKA (Windhoek, Namibia)}
\begin{center}
\begin{small}
\pgfplotstabletypeset{data/atmprof10.dat}
\end{small}
\end{center}
\end{document}
